Can COVID-19 Indicators Improve Forecasts and Hotspot Prediction?

### 2021-06-12 forecast/paper/paper.tex
Edits up through intro

% Please add a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about
  the significance of their research paper written at a level understandable to
  an undergraduate educated scientist outside their field of speciality. The
  primary goal of the significance statement is to explain the relevance of the
  work in broad context to a broad readership. The significance statement
  appears in the paper itself and is required for all research papers.}

@@

\dropcap{T}racking and forecasting indicators from public health reporting1
streams---such as confirmed cases and deaths in the COVID-19 pandemic---is
crucial for understanding disease spread, formulating public policy responses,
and anticipating future public health resource needs.  In a companion paper, we
describe our research group's (Delphi's) efforts in curating and maintaining a 
database of real-time indicators that track COVID-19 activity and other relevant
phenomena. The signals (a term we use synonomously with ``indicators'') in this
database are accessible through the COVIDcast API \cite{CovidcastAPI}, with
associated R \cite{CovidcastR} and Python \cite{CovidcastPy} packages for
convenient data fetching and processing tools. In the current paper, we aim to
quantify the additional utility provided by a core set of these indicators for
two fundamental prediction tasks: probabilistic forecasting of COVID-19 case
rates and prediction of future COVID-19 hotspots (where a hotspot is defined by
the event that the relative increase in COVID-19 cases in a given location
exceeds a certain threshold).

@@

In our companion paper, we analyze correlations between various indicators and   
COVID case rates. These correlations are natural summaries of the
contemporaneous association between an indicator and COVID cases, but they fall
short of delivering a satisfactory answer to the question that motivates the
current article: is the information contained in an indicator demonstrably
useful for the prediction tasks we care about? For such a question, lagged
correlations (e.g., measuring the correlation between an indicator and COVID
case rates several days in the future) move us in the right direction, but still
do not move us all the way there. The question about \emph{utility for
  prediction} is focused on a much higher standard than simply asking 
about correlations; to be useful in forecasting or hotspot prediction, an 
indicator must provide relevant information that is not otherwise contained in 
past values of the case rate itself. We address this in the most direct way
possible, by inpsecting the difference in predictive performance of simple 
autoregressive models trained with and without access to past values of a 
particular indicator.

There is a second and more subtle issue in evaluating predictive utility that
deserves explicit mention, as it will play a key role in our
analysis. Indicators computed from surveillance streams will often be subject to
latency and/or revision. For example, a signal based on aggregated medical
insurance claims may be available after just a few days, but it can then be
substantially revised over the next several weeks as additional claims are
submitted and processed late. Correlations between such a signal and case rates 
calculated ``after the fact'' (meaning, we look retrospectively and use the
finalized values of this indicator) will not deliver an honest answer to the
question of whether this indicator would have been useful in real time. Instead, 
we build predictive models using only the data that would have been available 
\emph{as of} the prediction date, and compare the ensuing predictions in terms
of accuracy. This leverages our group's {\tt evalcast} R package, which plugs
into the COVIDcast API's data versioning system, and facilitates honest
backtesting of forecasters.

Furthermore, adopting this practically-oriented forecasting
perspective to indicator evaluation requires one to confront
fundamental limitations inherent to public health surveillance
systems.  In particular, many reporting systems are plagued by
latency, with delays in reporting that can sometimes span
multiple weeks.  Latency manifests itself in different ways in different data sources.  Some data sources may only make a signal available one week after the date it describes; other
data sources may provide immediate imperfect initial estimates, which are
gradually ``backfilled'' over a period of days, weeks, or months (in which previous estimates are updated as new
data trickle in).  And few of
these latency patterns are stable or predictable, with holidays,
extreme weather events, and changes in the reporting process all
affecting the dynamics of the delay process.  To fully and accurately
understand an indicator's usefulness for forecasting requires
accounting for the unfortunate reality of data latency.

Examining the importance of additional features for prediction is a core question
in inferential statistics, and there are many ways to examine this question
under a variety of assumptions. For time series, work in
this area can be traced at least to Granger causality~\cite{GrangerCausality} and
related multivariate extensions~\cite{GewekeTimeSeries} under the condition 
that the series are linearly related autoregressions among other assumptions.
These techniques are especially popular in econometrics and neuroscience~\cite{mccracken2007asymptotics,NeuroscienceGranger}.

We choose to take a
predictive angle, which is simple but also fairly robust, compared to various
other classical inferential approaches. In fact, drawing rigorous inference based on
predictions, without (or with lean) assumptions, is an active field of
research~\cite{dai2021significance,fryer2021shapley,zhang2020floodgate,williamson2020unified}.
Nonetheless, the predictive approach directly examines the target of interest: do
these signals help to produce forecasts? Furthermore, directly comparing
forecast errors without strong assumptions through tests of predictive accuracy
are common in epidemiology, econometrics, and other
fields~\cite{DieboldMarianoTest,DieboldRetrospective}.

@@

% RJT: 
% What is a leading indicator?
% Conceptual vs empirical.

>>>>

% Please add a significance statement to explain the relevance of your work
% TODO write one. This is the indicator's one.
\significancestatement{To study the COVID-19 pandemic, its effects on society,
  and measures for reducing its spread, researchers need detailed data on the
  course of the pandemic. Standard public health data streams suffer
  inconsistent reporting and frequent, unexpected revisions. They also miss
  other aspects of a population's behavior, worthy of consideration. We
  present an open database of COVID signals in the United States, measured at
  the county level and updated daily. These include traditionally reported COVID
  cases and deaths, and many others: signals based on measures of mobility,
  social distancing, internet search trends, self-reported symptoms, and
  patterns COVID-related activity in de-identified medical insurance claims. The
  database provides all signals in a common, easy-to-use format, empowering both
  public health research and operational decision-making.
}

@@

\dropcap{T}racking and forecasting indicators from public health reporting1
streams---such as confirmed cases and deaths in the COVID-19 pandemic---is
crucial for understanding disease spread, formulating public policy responses,
and anticipating future public health resource needs.  In a companion paper, we
describe our research group's (Delphi's) efforts in curating and maintaining a 
database of real-time indicators that track COVID-19 activity and other relevant
phenomena. The signals (a term we use synonomously with ``indicators'') in this
database are accessible through the COVIDcast API \cite{CovidcastAPI}, with
associated R \cite{CovidcastR} and Python \cite{CovidcastPy} packages for
convenient data fetching and processing tools. In the current paper, we aim to
quantify the utility provided by a core set of these indicators for two
fundamental prediction tasks: probabilistic forecasting of COVID-19 case 
rates and prediction of future COVID-19 hotspots (where we define a hotspot by a
relative increase in COVID-19 cases, to be spelled out precisely later).

@@

In our companion paper, we analyze correlations between various indicators and   
COVID case rates. These correlations are natural summaries of the
contemporaneous association between an indicator and COVID cases, but they fall
short of delivering a satisfactory answer to the question that motivates the
current article: is the information contained in an indicator demonstrably
useful for the prediction tasks we care about? For such a question, lagged
correlations (e.g., measuring the correlation between an indicator and COVID
case rates several days in the future) move us in the right direction, but still
do not move us all the way there. The question about \emph{utility for
  prediction} is focused on a much higher standard than simply asking 
about correlations; to be useful in forecast or hotspot models, an indicator
must provide relevant information that is not otherwise contained in past values
of the case rate series itself. We will assess this in the most direct way
possible: by inpsecting the difference in predictive performance of simple
autoregressive models trained with and without access to past values of a
particular indicator.

There is another, more subtle issue in evaluating predictive utility that 
deserves explicit mention, as it will play a key role in our analysis.
Signals computed from surveillance streams will often be subject to  
latency and/or revision. For example, a signal based on aggregated medical
insurance claims may be available after just a few days, but it can then be
substantially revised over the next several weeks as additional claims are
submitted and/or processed late. Correlations between such a signal and case
rates calculated ``after the fact'' (computed retrospectively, using the
finalized values of this signal) will not deliver an honest answer to the    
question of whether this signal would have been useful in real time. Instead,
we build predictive models using only the data that would have been available
\emph{as of} the prediction date, and compare the ensuing predictions in terms
of accuracy. This leverages our group's {\tt evalcast} R package, which plugs
into the COVIDcast API's data versioning system, and facilitates honest
backtesting of forecasters.

% Furthermore, adopting this practically-oriented forecasting
% perspective to indicator evaluation requires one to confront
% fundamental limitations inherent to public health surveillance
% systems.  In particular, many reporting systems are plagued by
% latency, with delays in reporting that can sometimes span
% multiple weeks.  Latency manifests itself in different ways in different data sources.  Some data sources may only make a signal available one week after the date it describes; other
% data sources may provide immediate imperfect initial estimates, which are
% gradually ``backfilled'' over a period of days, weeks, or months (in which previous estimates are updated as new
% data trickle in).  And few of
% these latency patterns are stable or predictable, with holidays,
% extreme weather events, and changes in the reporting process all
% affecting the dynamics of the delay process.  To fully and accurately
% understand an indicator's usefulness for forecasting requires
% accounting for the unfortunate reality of data latency.

% Examining the importance of additional features for prediction is a core question
% in inferential statistics, and there are many ways to examine this question
% under a variety of assumptions. For time series, work in
% this area can be traced at least to Granger causality~\cite{GrangerCausality} and
% related multivariate extensions~\cite{GewekeTimeSeries} under the condition 
% that the series are linearly related autoregressions among other assumptions.
% These techniques are especially popular in econometrics and neuroscience~\cite{mccracken2007asymptotics,NeuroscienceGranger}.

% We choose to take a
% predictive angle, which is simple but also fairly robust, compared to various
% other classical inferential approaches. In fact, drawing rigorous inference based on
% predictions, without (or with lean) assumptions, is an active field of
% research~\cite{dai2021significance,fryer2021shapley,zhang2020floodgate,williamson2020unified}.
% Nonetheless, the predictive approach directly examines the target of interest: do
% these signals help to produce forecasts? Furthermore, directly comparing
% forecast errors without strong assumptions through tests of predictive accuracy
% are common in epidemiology, econometrics, and other
% fields~\cite{DieboldMarianoTest,DieboldRetrospective}.

@@
